Low Complexity Adaptive Sonar Imaging

### 2015-10-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:21	14:23:06	(Wed21.Oct)

\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?

@@

\IEEEPARstart{T}{he} best adaptive sonars are found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated to post process the received data. Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

@@

In this study we apply LCA to active sonar imaging. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.

Our results show that MVDR, when adapted to work in active sonar imaging, computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. A conclusion is provided in \ref{sec:conclusion}.

@@

We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows. However, it needs to be able to select between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity. We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$. For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters. Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively. Hence, it can be used to control aggressiveness and resolution gain of LCA.

In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR. Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using GPUs. Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.

@@

Throughout this article we have used LCA with the Kaiser window function. However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function. For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. To derive it, we start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
\end{align}
%
where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
%
\begin{align}
\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
Unity gain in the look direction is ensured as long as the weights sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
\end{align}
%
This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.

>>>>

\newif\ifPeerReview\PeerReviewfalse              % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?

@@

\IEEEPARstart{T}{he} best adaptive sonars are found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

@@

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Section \ref{sec:conclusion} provides a conclusion.

@@

We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows. However, it needs to be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity. We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$. For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters. Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively. Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.

In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR. Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g. GPUs. Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.

@@

Throughout this article we have used LCA with the Kaiser window function. However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function. For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. The derivation generalizes the one described for a non-steered window in~\cite{Stoica2005} to also apply for the steered (complex) window. We start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
\end{align}
%
where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
%
\begin{align}
\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
Unity gain in the look direction is ensured as long as the weights sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
\end{align}
%
This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
